Strict Pure Functional Programming and Automatic Parallelization\\ (New-project-in-progess Paper)

### 2016-03-12 pubs/tfp16/paper.tex
synching

\begin{abstract}
The trend is functional programming in scientific computing, and in particular
strict (by default) functional programming, in various guises.  Our project is
the demonstration of a light-weight, higher-order, polymorphic pure functional
language implementation in which we can experiment with automatic
parallelization strategies and varying degrees of default strictness.
Strictness prescribes the degree of potential parallelism, that is, we are not
exploring speculative evaluation.

@@

\section{Background and Motivation}
The practice of highest-performance scientific computing (HPC) is, overall,
highly conservative with respect to change.  Even as the US government commits
to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
highly respected member of the HPC community recently stated ``Fortran is
essential for exascale programming.''~\cite{Heroux16} Existing US government
laboratory HPC codes run to hundreds of thousands of lines and complete
rewrites are infeasible.\footnote{A European consortium is anticipating
  exascale computing as well~\cite{EUexascale}, but the authors are not
  knowledgeable about the state of large-scale, non-USA HPC codes.}  C++ has
become more common for new code starts, but their nature remains the same:
huge codes with fair to non-existent encapsulation of side effects, and
multiple levels of parallelism also poorly abstracted, especially at the
thread (shared-memory) level.  Thread-level parallel efficiency can be poor
and is explained by Amdahl's law because the parallel model is bulk
synchronous.

@@

While our interpretation of STG is a higher-order, polymorphic, pure
functional language, some syntactic restrictions make direct programming
rather tedious.  While a near-term goal is to bridge the gap\ldots

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER, the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students, and the DOE Science Undergraduate Laboratory Internship program for
undergraduate students.
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).

>>>>

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Strictness constrains
the degree of potential parallelism, that is, we are not exploring speculative
evaluation.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade, a highly respected member of the HPC
community recently stated publicly that ``Fortran is essential for exascale
programming.''~\cite{Heroux16}\footnote{Various EU, Japanese, and
  international groups are also anticipating exascale
  computing.~\cite{EUexascale,Exascale-org}} Existing US government laboratory
HPC codes run to hundreds of thousands of lines of Fortran and complete
rewrites are infeasible.~\footnote{The codes of our lab's British counterpart
  are reportedly in the same state.}  C++ has become more common for new code
starts, but their nature remains the same: huge codes with fair to
non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

@@

As mentioned, a current near-term goal is to use ghc as our front-end
to generate STG as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented
a \emph{mini-Haskell} front-end, a strict subset of standard Haskell syntax.

???give a proper grammar and summarize any tricky translations???

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER, the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students, and the DOE Science Undergraduate Laboratory Internship program for
undergraduate students.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).
